---
title: "Filter and Arrange"
subtitle: Filter rows and arrange them in a specific order
format:
    html:
        df-print: paged
---

::: {.callout-note title="Required packages and data for this tutorial"}

In this tutorial we use the `tidyverse` package and the simulated [practice data](../../data-management/index.qmd#practice-data).

```{r, message=FALSE}
library(tidyverse)
d <- read_csv("https://data.dataliteracy.cc/simulated/practice_data.csv")
```
:::

# Subsetting rows with filter() {#filter} 

The `filter` function can be used to select a subset of rows. 
The first argument of the `filter` function is the tibble you want to filter.
The second argument is the condition that should be met for a row to be included in the result.
Let's say we want to select only the rows where the `experiment_group` column is equal to `control`.
We can then use the `==` (is equal to) operator.

```{r}
filter(d, experiment_group == 'control')
```

This gives us the 200 rows (out of 600) where the `experiment_group` is `control`.

We can use other common operators as well, such as `>`, `<`, `>=`, `<=`, and `!=` (not equal).
And we can also combine multiple conditions with the `&` (and) and `|` (or) operators. 

For example, here we select all rows where the `experiment_group` is `control` and the `age` is greater than 30:

```{r, output=FALSE}
filter(d, experiment_group == 'control' & age > 30)
```

A less common operator that is usefull to know about is `%in%`, which is used to check if a value is in a list of values.
For example, to select all rows where the `experiment_group` is either `positive` or `negative`:

```{r, output=FALSE}
filter(d, experiment_group %in% c('positive', 'negative'))
```

You an also *invert* any condition by putting a `!` (NOT) in front of it.
So the following code selects all rows where the `experiment_group` is NOT in the list `positive` or `negative`:

```{r, output=FALSE}
filter(d, !experiment_group %in% c('positive', 'negative'))
```

::: {.callout-note title="A deeper understanding of the filter condition" collapse="true"}

Based on the examples given you probably already have a good enough understanding of how the `filter` function works to use it in your own code.
In this optional information block we'll go a bit deeper into how the filter condition works, and what operators you can use.

## The condition is a logical expression

The condition in `filter` can be any **logical expression**.
A logical expression is simply a statement that is either `TRUE` or `FALSE`.
When we use a logical expression in the `filter` function, we are asking R to evaluate this expression for each row in the tibble. 
Each row for which the expression evaluates to `TRUE` is then included in the subset.

If you know a bit about how logical expressions work, you will have great control over what rows are included in your subset.
Here is an overview of the most important operators for logical expressions.

### Comparison operators

Comparison operators are used to compare two values.

- `==`   equal to
- `!=`   not equal to
- `>`    greater than
- `>=`   greater than or equal to
- `<`    less than
- `<=`   less than or equal to
- `%in%` is in a list of values (second value must be a list or vector) 

Example: 

```{r}
5 > 1    # TRUE:  5 is greater than 1
5 < 1    # FALSE: 5 is less than 1
"A" %in% c("A", "B", "C")  # TRUE: "A" is in the list
"A" %in% c("B", "C", "D")  # FALSE: "A" is not in the list
```

### Logical operators

Logical operators are used to combine multiple conditions.

- `&` and
- `|` or
- `!` not

Example:
```{r}
5 > 1 | 5 < 1   # TRUE: 5 is greater than 1 OR 5 is less than 1
5 > 1 & 5 < 1   # FALSE: 5 is greater than 1 AND 5 is less than 1
!5 < 1          # TRUE: it is not the case that 5 is smaller than 1
```

### Using equations

You can also use equations in your conditions.
For example, to select all rows where the absolute difference between `trust_t1` and `trust_t2` is greater than 2:

```{r, output=FALSE}
filter(d, abs(trust_t2 - trust_t1) > 2)
```

### Parentheses

For complex conditions, you can use parentheses to group conditions, similar to how you would in a mathematical expression.
For example, say that you want to inspect *surprising* cases where trust in journalists decreased after watching the positive movie, or increased after watching the negative movie.

```{r, output=FALSE}
filter(d, (experiment_group == 'positive' & trust_t2 < trust_t1) |
          (experiment_group == 'negative' & trust_t2 > trust_t1))
```
:::

# Filtering out missing values

Filtering out cases with missing values works a bit differently.
Missing values in R are represented by `NA`, but you **cannot (!!)** use something like `filter(d, age != NA)`.
Instead, you can use the `is.na` function to check if a value is missing.

```{r}
filter(d, is.na(age))     ## rows where age IS missing
```
```{r, results='hide'}
filter(d, !is.na(age))    ## rows where age IS NOT missing
```

In addition, there is a special function in the `tidyverse` for removing rows if ANY column in the data is missing.
This is the `drop_na` function.

```{r, results='hide'}
drop_na(d)
```

This returns 595 rows (out of 600) because 5 rows had missing values in the `age` column.

# Sorting rows with arrange() {#arrange}

The `arrange` function can be used to sort the rows of a tibble.
For example, if we want to sort the data by the `age` column:

```{r}
arrange(d, age)
```

By default, the rows are sorted in ascending order.
If you want to sort in descending order, you can put a minus in front of the variable name.

```{r, output=FALSE}
arrange(d, -age)
```

If you want to sort on multiple columns, you can simply add them to the `arrange` function.
For example, to sort first on `experiment_group` (ascending) and then on `age` (descending):

```{r, output=FALSE}
arrange(d, experiment_group, -age)
```


