
---
title: "Mutate and recode"
subtitle: "Create new variables and recode existing ones"
order: 5
format:
  html:
    df-print: paged
---

::: {.callout-note title="Required packages and data for this tutorial"}

In this tutorial we use the `tidyverse` package and the simulated [practice data](../../data-management/index.qmd#practice-data).

```{r, message=FALSE}
library(tidyverse)
d <- read_csv("https://data.dataliteracy.cc/simulated/practice_data.csv")
```
:::

# Creating and modifying columns with mutate() {#mutate}

The `mutate` function allows you to create or modify columns in a data frame.
You just specify which columns you want to create or modify, and then provide an expression for how to compute the new values.
In this expression you can refer to other columns in the table, and use any R functions you like, which makes `mutate` a very powerful tool.

## Creating new variables

To create a new variable you use the following syntax:

```{r, eval=F}
mutate(d, new_variable = expression)
```

The expression can be anything that returns a valid column.
For example, in the practice data we have the columns `trust_t1` and `trust_t2`, which represent trust in journalists before and after the experiment.
We can create a new variable `trust_change` that represents the change in trust from before to after the experiment.

```{r}
d <- mutate(d, trust_change = trust_t2 - trust_t1)

select(d, trust_t1, trust_t2, trust_change)
```


## Mutate existing variables

To mutate an existing variable, you can simply overwrite the column with the same name.
For example, let's say that we want to standardize the `trust_change` variable that we just made.
We can standardize a variable with the `scale` function, so we can use that inside of `mutate`.

```{r}
d <- mutate(d, trust_change = scale(trust_change))
```

Now the `trust_change` variable is standardized, which means that it has a mean of 0 and a standard deviation of 1.
A nice way to get a quick overview of the distribution of a single variable is to plot a histogram.

```{r, results=T}
hist(d$trust_change)
```


# Recoding variables {#recode}

Recoding variables means changing the values of a variable based on some condition.
This is a common operation in data management, because often you want to change the values of a variable to make them more interpretable, correct errors, or prepare the data for analysis.
To recode variables in R, you can use the `mutate` function in combination with the `case_match` and `case_when` functions.



## Recode with `case_match`

The `case_match` function is a simple way to recode specific values into new values.
For example, in our practice data we have a column with the experimental groups, which are `control`, `positive`, and `negative`.
Let's say we want to clarify that `positive` means `positive_movie`, and `negative` means `negative_movie`.
We could then use `case_match` to change these values.

```{r}
d <- mutate(d, experiment_group = case_match(experiment_group,
                                            "positive" ~ "positive_movie",
                                            "negative" ~ "negative_movie",
                                            .default = experiment_group))
```


Here we say: overwrite the `experiment_group` column with output of the `case_match` function.
Inside the `case_match` function, we specify three things:

1. The column we want to recode (`experiment_group`).
2. The conditions for recoding the values. We have two conditions:
   1. If value is `"positive"`, recode into `"positive_movie"`.
   2. If value is `"negative"`, recode into `"negative_movie"`.
3. We specify a `.default` value for values that are not matched in the conditions. Here we say that in that case we want to use the current value of the `experiment_group` column.

If you check the unique values of the `experiment_group` column, you will see that `positive` and `negative` have been changed to `positive_movie` and `negative_movie`, and that `control` remains the same.
```{r}
unique(d$experiment_group)
```

## More flexible recoding with `case_when`

The `case_match` function is great if you need to recode many values, but sometimes you need more flexibility.
For example, if we want to recode the `age` variable into categories (e.g., `<= 20`, `20-30`), it would be really tiresome to recode every individual age value.
With the `case_when` function, we can specify the conditions using **logical expressions**.
Each condition is evaluated in order, and the first one that is `TRUE` is used.

```{r}
d <- mutate(d, age_category = case_when(
  age < 20 ~ "<= 20",
  age < 30 ~ "20-30",
  age < 40 ~ "30-40",
  age < 50 ~ "40-50",
  age < 60 ~ "50-60",
  .default = ">= 60"
))

table(d$age_category)
```

## Binary cases with `if_else`

If you only have two categories, you can use the `if_else` function.
You could technically also use `case_when` for this, but `if_else` is more concise and easier to read.
The syntax for `if_else` is:

```{r, eval=F}
if_else(condition, value_if_true, value_if_false)
```

A common use case is that sometimes you want to perform an operation only on a subset of the data.
For example, in our data there are a few participants that accidentally entered their birthyear instead of their age. 
To correct this, we can use `if_else` to set the age to `2024 - birthyear`, but only if the number the participants entered is above 1000 (which is only the case if it's a birthyear).

```{r}
d <- mutate(d, age = if_else(age > 1000, 2024 - age, age))
```

So this reads: if the age is above 1000, return `2024 - age`, otherwise return the current age.
