---
title: "Using the pipe syntax"
subtitle: "How to make your code more readable"
order: 6
editor_options: 
  chunk_output_type: console
---


# Working with Pipes

In the [Functions](../getting-started/functions.qmd#using-the-pipe-syntax) tutorial we already mentioned that R also has a `pipe` syntax. 
This is a way to chain functions together, where the output of one function is the input of the next function.
The syntax for the pipe is `|>`, and it is used like this:

```{r, eval=F}
argument1 |> function(argument2)
```

The first argument of a funtion can also be *piped into it*. 
Between the parentheses of the function we then only need to specify any additional arguments (if needed).
Often the first argument of a function is the input data, which allows you to chain functions together.

```{r, eval=F}
data |> 
   do_this() |> 
   then_this() |>
   finally_this()
```


## Using pipes with the tidyverse

The tidyverse is designed to work really well with pipes.
All of the functions for working with a tibble (like `select`, `filter`, `arrange`, etc.) have the first argument as the tibble itself, and the output is also a tibble.
This means that you can chain these functions together to create a single pipeline for cleaning and preparing your data.

For example, the following code reads a csv file, selects columns, filters rows and finally arranges the data:

```{r, message=F}
library(tidyverse)

practice_data = read_csv("https://data.dataliteracy.cc/simulated/practice_data.csv")
practice_data = select(practice_data, age, experiment_group, trust_t1)
practice_data = filter(practice_data, age >= 18)
practice_data = arrange(practice_data, trust_t1)
```

With pipes, we can write the same code in a more readable way:

```{r, message=F}
practice_data = read_csv("https://data.dataliteracy.cc/simulated/practice_data.csv") %>%
  select(age, experiment_group, trust_t1) %>%
  filter(age >= 18) %>%
  arrange(trust_t1)
```

::: {.callout-note title="The alternative pipe symbol `%>%`" collapse="true"}


There is another pipe symbol in R: `%>%`.
In this book we will always use `|>`, but it's good to know about the existence of `%>%`, because you might encounter it in other resources.

Both functions work almost in the same way.
So why have two?
The reason is simply that R keeps evolving, and the `|>` pipe was only recently introduced.

The `%>%` pipe was introduced in the `magrittr` package, and was made popular by the `tidyverse`. 
Because of this popularity, R decided that it would be a good idea to have a *native* pipe in the language itself, meaning that you don't need to install a package to use it. 
This is why they introduced the `|>` pipe.
:::