---
title: Read and write
order: 1
subtitle: How to get data into R, and how to store it
---


# Reading and writing data {#importing-data}

R can read files from many types of file formats.
Here we will focus on the **csv** format, which is one of the most common formats for storing and sharing rectangular data (i.e., data in rows and columns).
Once you know how to read a CSV file, you can easily read other file formats as well (e.g., **Excel**, **SPSS**).

## CSV files

::: {.callout-note title="What is a CSV file?" collapse="true"}
`CSV` stands for Comma Separated Values.
It is a simple text file, that you can open in any text editor.
In order to store a data frame (i.e. data in rows and colums), it simply read every line as a row, and separates the columns by a comma (or sometimes another symbol, like a semicolon).

For example, the following CSV file contains a data frame with three columns: `resp_id`, `gender`, and `height`. 
The first row contains the column names, and the following rows contain the data.

```{r, eval=F}
resp_id,gender,height
1,M,176
2,M,165
3,F,172
4,F,160
```

The benefit of this simplicity is that any respectable spreadsheet or statistical software (e.g., Excel, Google sheets, SPSS, Stata) can read it. 
This makes CSV files a great way to share and store data.

And just in case you're worried, yes, CSV can also handle textual data.
It uses some tricks to make sure that commas inside the text are not interpreted as column separators.
:::


To show you how to work with CSV files, we'll first import a dataset from the internet. Then we'll show how to write this data to a CSV file on your computer, and how to read it back into R.

### Importing data from a URL

To read CSV files into R, you can use the `read_csv` from the tidyverse (more specifically from the `readr` package).
If you provide a URL, it will download the file from the internet.
Here we read the data and assign it to the name `d` (short for data).
You can use any name you like, but since you'll be referring to this data a lot, it's convenient to keep it short.

```{r, message=FALSE}
library(tidyverse)

d <- read_csv("https://data.dataliteracy.cc/simulated/practice_data.csv")
```

Make sure to always check whether the data was imported correctly:

```{r, results=FALSE}
d
```

You can also view the data in a larger spreadsheet-like view using the `View` function.
Either click on the name (`d`) in the **Environment** tab in RStudio (top right panel), or use the `View` function:

```{r}
View(d)

```

This will open a new tab in RStudio that shows all the data. 
In the top menu bar you can also `filter` the data and `search` for specific values, or click on column names to sort the data.


### Writing data to a CSV file on your computer

You can use the `write_csv` function to write a tibble to a CSV file on your computer.
If you just provide a file name, it will be saved in your current [working directory](../getting-started/file-management.qmd#working-directory).

```{r, eval=FALSE}
write_csv(d, "practice_data.csv")
```


### Reading data from a CSV file on your computer

Now let's read this file back into R.
Since the file is in your working directory, you can just specify the file name:

```{r, eval=FALSE}
d2 <- read_csv("practice_data.csv")
```

You can check and verify that the data (d2) is indeed identical to the original data (d).

::: {.callout-warning title="CSV pitfalls to avoid" collapse="true"}
There are two important pitfalls to avoid when working with CSV files:

### Pitfall 1: Corrupting the file by opening it in Excel

When you download a CSV file from the internet, some computers might immediately ask you whether you want to open it in your default spreadsheet program (e.g., Excel, Numbers). 
Do not do this, but instead download the file directly to your computer. If you open the file and accidentally save it, it can overwrite the CSV file with a different format. 
Excel in particular has a habit of breaking CSV files this way. 

### Pitfall 2: Different flavours of CSV files

There are different flavours of CSV files (for historic reasons). 
Even though we call them "comma separated values", the separator is sometimes a semicolon or a tab. 
And depending on language, the decimal separator can be a comma or a dot. In particular, there are two most common versions of the CSV file. 
This is why tidyverse has two `read_csv` functions: `read_csv` and `read_csv2`. In general, you can just try `read_csv` first, and if it doesn't work, try `read_csv2`.
:::

## Reading other file formats, like Excel and SPSS

Now that you know how to read and write CSV files, reading other file formats is a piece of cake.
It works almost the same way, but you just need to download a package that can read the file format.

For instance, to read an Excel file, you can use the `readxl` package, which provides the `read_excel` function. 
To read an SPSS file, you can use the `haven` package, which provides the `read_sav` function.
You might have to take care of some additional details, such as the sheet name in the Excel file, or the variable labels in the SPSS file.
But once you've got the hang of managing your data with the tidyverse, you'll be able to handle any data frames formats that come your way.

