---
title: Names and Objects
subtitle: Remember information by giving it a name
order: 2
---

# The power of assignment 

In R, and in computer programming in general, the most essential operation is to assign **objects** to **names**.
By **object**, we then broadly mean any piece of information. a single number, a text, a list of numbers, and even an entire data set. 

In plain terms, **assignment** is how you make R remember things by assigning them to a name. To assign an object to a name, we use the **arrow notation**: `name <- value`. For example:

```{r}
x <- 2
```

::: {.callout-note title="Good to know: you can also use `x = 2`" collapse="true"}
Instead of using the **arrow** notation, you can also use the **equal sign** notation: `name = object`.

```{r}
x = 2
```

We will in general always use the **arrow** notation. But if you encounter the **equal sign** notation, just remember that it's the same thing.
:::

By **running** the code `x <- 2`, you are saying: **Assign the value `2` to the name `x`**. 
Any objects that you assigned to names are stored in your **Environment**. You can see this environment in the top-right window, under the **Environment** tab.
If you assigned `2` to `x`, you should see a table with in the left column the names (`x`) and in the right column a description of the object. 
For simply objects like numbers, this will just be the value (`2`).

From hereon, when you use the name `x` in your code, it will **refer** to the value `2`. So when we run the code `x * 5` (x times 5) it will print the number `10`

```{r}
x * 5
```

::: {.callout-note title="Why does R print a [1] before the result?" collapse="true"}
When running `x * 5`, R correctly prints the value 10. But why does it say `[1] 10`?
The reason is that R always thinks of a number (or string) as a **vector** (i.e. list of values), that can have 1 or multiple values. 
The `[1]` indicates that 10 is the first (and only) value.

If you print a longer **vector**, you can see that R prints `[...]` at the start of each line, just to help you see the position of individual values.
The following code generates a vector with numbers from 1 to 50

```{r}
1:50
```
:::



## Assigning versus printing

Notice that when you ran the code `x <- 2`, R didn't print any values to the console (the bottom-left window).
But when you ran `x * 5`, R did print the value `10`.
Basically, when you run code, and you DO NOT assign the result to a name, R will print the result to the console.

So the following code will NOT print the object (`"I will not be printed"`) to the console, but will store it (you can see the name pop up in the **Environment** tab)

```{r}
i_will_be_remembered <- "I will not be printed"
```

And the following object (`"I will be printed"`) will be printed to the console, but not stored in the **Environment**.

```{r}
"I will be printed"
```



## Assigning different types of objects

You can assign any type of object to a name, and you can use any name, as long as it starts with a letter and doesn't contain spaces or symbols (but underscores are OK)

```{r}
a_number = 5
my_cats_name = "Hobbes"
```

If you run this code and check you **Environment** (top-right), you should now see these name-object pairs added.


## Assigning results 

Till now we only directly assigned objects to names. This is convenient, but the power of assignment really shines when you use it to store **results**. 
For example, we can also do this.

```{r}
x = 5 + 10
```

This a very simple example, but just think for a second what this allows us to do. 
Since we can assign anything to a name, we can break down any complicated procedure into multiple steps! 
For now, the key lesson is just to wrap your head around the syntax for assigning objects to names. 
This is fundamental to everything you will be doing in R (and in programming in general). 




