---
title: R in Action (demo)
---

R is a very powerful tool, but it takes some time to learn how to use it before you get to fully appreciate what you can use it for.
On this page we show you a **quick example of some of the things you will learn here**.

All the code that you see here will be explained in the rest of this online book. 
**For now, don't worry about understanding the code**, and focus on thinking how the techniques we'll be using fit into your tool belt as a communication scientist.

## Loading the packages we'll use

One of the things that makes R so versatile, is that anyone can extend it by writing new *packages*. 
You can think of packages kind of like apps in an app-store. 
For this demo, we'll need two packages, that you'll first need to install.

```{r, eval=F}
install.packages('tidyverse')
install.packages('sjPlot')
```

You only need to install packages once, just like apps on you mobile phone.
Once downloaded, they are stored in your R *library*.
When you then use the packages in an R script, you just *open* them like this:

```{r, message=F}
library(tidyverse)
library(sjPlot)
```

## Example 1: Importing and cleaning data

The first step for any analysis is to import data and clean it up.
In the [Data Management](../data-management/index.qmd) chapter, you will learn how to use the `tidyverse` and `pipe` syntax to create clear and transparent data cleaning pipelines.

In this example we show a pipeline that performs the following steps:

* Import a data set from the internet
* Select the columns that we need, and rename one inconvenient name
* Correct that some participants entered their birth year instead of their age
* Filter out rows where the age is below 18 or missing (`NA`)

```{r}
library(tidyverse)

d <- read_csv("https://data.dataliteracy.cc/simulated/practice_data.csv") |>
  select(age, np_subscription, news_consumption = `news consumption`) |>
  mutate(age = ifelse(age > 200, 2024 - age, age)) |>
  filter(!is.na(age) & age >= 18)
```

## Example 2: Visualizing data

```{r}
ggplot(d, aes(x = age, y = news_consumption, color = np_subscription)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Trust in T1 by age",
       x = "Age",
       y = "Trust in T1")
```


## Import data

The first step for any analysis is to import your data. 
Using the `read_csv` function from the `tidyverse` package, we can directly download this information from the internet and import it into R.




# TO DO

We'll show a nice example here of how tidyverse allows us to import and clean a data set nice and quick, how we can visualize it, perform a statistical test, and create an APA ready table and visualizations
