---
title: "5.3 Statistical Power & Sample Size"
subtitle: "Planning sample sizes and avoiding underpowered studies with the pwr package"
---

# Introduction to Statistical Power

When designing an empirical study, one of the most critical decisions is **how many participants to recruit**.
If your sample is too small, your study may be **underpowered**: even if a real effect exists in the population, your test may have a low probability of detecting it.

Statistical power is defined as:

$$\text{Power} = 1 - \beta = P(\text{Reject } H_0 \mid H_1 \text{ is true})$$

In communication science, the conventional standard is to aim for a power of at least **0.80** (an 80% chance of detecting an effect if it truly exists).

# The Four Interconnected Parameters

Statistical power involves a four-way relationship between:

1. **Significance level ($\alpha$)**: Usually set to 0.05.
2. **Statistical power ($1 - \beta$)**: Usually targeted at 0.80.
3. **Effect size**: The expected magnitude of the phenomenon in the population (e.g., Cohen's $d$, Pearson's $r$, or Cohen's $f^2$).
4. **Sample size ($N$)**: The number of observations needed.

Given any three of these parameters, the fourth can be calculated mathematically.

# Power Analysis for Independent Samples T-Test

::: {.callout-note title="Required Packages"}
```{r, message=FALSE}
library(pwr)
```
:::

Suppose we are planning an experiment comparing two independent groups (e.g., exposure to an interactive health website vs. a static text website).
We expect a moderate effect size of Cohen's $d = 0.50$, and want $80\%$ power at $\alpha = 0.05$.

We can compute the required sample size per group using `pwr.t.test()`:

```{r}
pwr.t.test(d = 0.50, sig.level = 0.05, power = 0.80, type = "two.sample", alternative = "two.sided")
```

The output indicates that we need approximately **64 participants per group**, or a total sample of $N = 128$.

## Visualizing Power Curves

How does statistical power change as sample size increases? We can generate and plot a power curve:

```{r}
library(tidyverse)

sample_sizes <- seq(10, 150, by = 5)
powers <- sapply(sample_sizes, function(n) {
  pwr.t.test(n = n, d = 0.50, sig.level = 0.05)$power
})

power_df <- tibble(n_per_group = sample_sizes, power = powers)

ggplot(power_df, aes(x = n_per_group, y = power)) +
  geom_line(color = "#0d9488", linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  labs(
    title = "Power Curve for Two-Sample T-Test (d = 0.50)",
    x = "Sample Size per Group (n)",
    y = "Statistical Power (1 - Beta)"
  ) +
  theme_minimal()
```

# Power Analysis for Regression

For multiple regression models, effect sizes are expressed as Cohen's $f^2$:

$$f^2 = \frac{R^2}{1 - R^2}$$

For instance, an $R^2$ of $0.09$ corresponds to $f^2 \approx 0.10$ (a small-to-moderate effect).
Using `pwr.f2.test()`:

```{r}
# u = number of predictors (e.g. 3 predictors), v = degrees of freedom for error
pwr.f2.test(u = 3, f2 = 0.10, sig.level = 0.05, power = 0.80)
```

The total sample size required is $N = v + u + 1 \approx 109 + 3 + 1 = 113$ observations.

# Best Practices for Theses and Research Projects

1. **Conduct A Priori Power Analyses**: Always perform power planning *before* data collection begins.
2. **Base Effect Sizes on Theory or Meta-Analyses**: Avoid post-hoc power calculations based on observed $p$-values.
3. **Plan for Attrition**: In survey panels or longitudinal designs, recruit 10–20% more participants to account for dropouts and incomplete responses.
