---
title: "8.1 Content Analysis & Text as Data"
subtitle: "An overview of computational text analysis in communication science"
---

# Introduction to Computational Content Analysis

Communication science is fundamentally about messages, narratives, and interactions.
For decades, researchers analyzed text through manual content analysis: developing codebooks, training human coders, and calculating inter-coder reliability.

Today, digital platforms and digitized archives produce text data at scales impossible for manual coding alone.
**Automated Content Analysis** allows researchers to study large collections of text systematically.

> **Course Companion:**
> For an in-depth textbook treatment of computational text analysis with full code tutorials, please consult [R for Communication Science](https://r4css.vanatteveldt.com/) by Wouter van Atteveldt et al. This module provides the conceptual canon and foundational tidyverse workflows expected in our methods curriculum.

# The Text as Data Pipeline

A computational text analysis project typically follows four steps:

1. **Data Acquisition**: Collecting texts via APIs, RSS feeds, digital archives, or web scraping.
2. **Text Representation (Tokenization)**: Converting unstructured character strings into structured rectangular data.
3. **Measurement / Classification**: Extracting meaning using dictionaries, supervised machine learning, or Large Language Models (LLMs).
4. **Statistical Analysis**: Modeling the extracted features as variables alongside survey or behavioral data.

# Tidytext: Tokenization & Word Frequencies

In R, the `tidytext` package follows the tidy data philosophy by representing text in a one-token-per-row format.

::: {.callout-note title="Required Packages"}
```{r, message=FALSE}
library(tidyverse)
library(tidytext)
```
:::

Consider a small sample of political statements:

```{r}
statements <- tibble(
  doc_id = 1:3,
  text = c(
    "We must invest in public education and clean energy.",
    "Economic growth requires lowering taxes and reducing regulation.",
    "Education reform and economic stability are top priorities."
  )
)
```

## Tokenizing: `unnest_tokens()`

The function `unnest_tokens()` splits each text into individual words, converts them to lowercase, and strips punctuation:

```{r}
tokens <- statements |> 
  unnest_tokens(word, text)
tokens
```

## Removing Stop Words

Many of the most frequent words in language are function words (e.g., "in", "and", "the", "are") that carry little topical information. `tidytext` provides a built-in `stop_words` dataset that we can remove using `anti_join()`:

```{r}
data("stop_words")

tidy_tokens <- tokens |> 
  anti_join(stop_words, by = "word")
tidy_tokens
```

## Word Frequency Counts

Once in tidy format, we can use standard `dplyr` and `ggplot2` commands to inspect term frequencies:

```{r}
tidy_tokens |> 
  count(word, sort = TRUE)
```

# Measurement Approaches

Depending on your research question, three main computational approaches are common:

## 1. Dictionary-Based Analysis
A **sentiment dictionary** or substantive lexicon maps specific words to categories (e.g., positive vs. negative tone, moral values, emotional expressions). 
By joining tokens with a lexicon (such as `get_sentiments("bing")`), each document can be assigned a tone score. While transparent and fast, dictionaries can struggle with context, sarcasm, and domain-specific meanings.

## 2. Supervised Machine Learning
When labels are needed for nuanced concepts (such as framing, toxicity, or policy topic), human coders first annotate a subset of texts (a training set). Algorithms (e.g., Naive Bayes, Support Vector Machines) learn to predict those labels on the remaining unlabelled dataset.

## 3. Large Language Models (LLMs)
Modern generative models (e.g., OpenAI GPT, local models via Ollama) can perform zero-shot and few-shot annotation by following natural language instructions. While highly capable of understanding context, researchers must rigorously validate LLM annotations against gold-standard human benchmarks.

# Next Steps & Advanced Reading

- **In-Depth Tutorials**: See [R for Communication Science: Automated Text Analysis](https://r4css.vanatteveldt.com/).
- **Advanced Techniques**: Topic modeling (LDA/STM) and word embeddings are covered in our elective computational trajectories.
