---
title: "9.1 Advanced Modeling Overview"
subtitle: "An optional guide to multilevel models, generalized linear models, and structural equations"
---

# Advanced Statistical Modeling

In undergraduate communication science methods, the linear model (OLS regression, ANOVA, and ANCOVA) forms our analytical bedrock.
However, empirical communication data often violates standard OLS assumptions:
- Categorical or binary outcomes (e.g., voting choice, clicking an ad).
- Hierarchical / clustered observations (e.g., students nested within classrooms, respondents within countries, repeated measures over time).
- Complex latent measurement networks with simultaneous direct and indirect pathways.

This optional reference summarizes the primary advanced modeling frameworks taught in our graduate and honors curricula.

::: {.callout-note title="Elective / Advanced Material"}
These models are not required for standard undergraduate assignments. They serve as an analytical roadmap for master's theses and research internships.
:::

# 1. Generalized Linear Models (GLMs) & Logistic Regression

When the dependent variable is binary (0/1) rather than continuous, OLS regression can produce nonsensical predictions outside $[0, 1]$.
Logistic regression transforms the outcome via the logit link function:

$$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \dots$$

In R, GLMs are fitted using the built-in `glm()` function:

```{r, eval=FALSE}
# Example logistic regression
model_logit <- glm(voted ~ age + news_exposure, data = survey_data, family = binomial(link = "logit"))
summary(model_logit)
```

# 2. Multilevel / Mixed-Effects Models (HLM)

When observations are grouped (e.g., participants nested in schools or longitudinal panel data), standard errors in OLS are underestimated because observations within clusters are not independent.

**Multilevel models** partition variance into within-group and between-group components.
In R, the standard package is `lme4`:

```{r, eval=FALSE}
library(lme4)

# Random intercept model: respondents nested within countries
model_mlm <- lmer(trust ~ news_use + (1 | country), data = comparative_data)
summary(model_mlm)
```

# 3. Structural Equation Modeling (SEM) & Confirmatory Factor Analysis (CFA)

While Module 3.2 uses Exploratory Factor Analysis (EFA), **Confirmatory Factor Analysis (CFA)** tests whether an explicit, theoretically specified factor structure fits empirical data.
Furthermore, SEM allows testing entire causal networks of latent variables and observed indicators simultaneously.

In R, the leading package is `lavaan`:

```{r, eval=FALSE}
library(lavaan)

cfa_model <- '
  # Measurement model
  trust =~ trust_fair + trust_unbiased + trust_accurate
  # Structural regression
  trust ~ news_exposure + political_interest
'

fit <- sem(cfa_model, data = survey_data)
summary(fit, fit.measures = TRUE, standardized = TRUE)
```

# Recommended Learning Resources

- **Gelman & Hill (2007)**: *Data Analysis Using Regression and Multilevel/Hierarchical Models*.
- **Rosseel (2012)**: *lavaan: An R Package for Structural Equation Modeling*.
- [R for Communication Science](https://r4css.vanatteveldt.com/)
