---
title: "Working with R"
format:
  html:
    number-sections: true
    number-offset: [8]
---

## Installation

To work with R, you will need to install two pieces of software.

* **R**. This is the actual R software, that is used to run R code.
* **RStudio**. This is a graphical user interface (GUI) that makes working with R much easier and prettier.

Both programs can be downloaded for free, and are available for all main operating systems (Windows, macOS and Linux).

### Installing R

To install R, you can download it from the [CRAN (comprehensive R Archive Network) website](https://cran.r-project.org/). Do not be alarmed by the website's 90's aesthetics. The website is legit.

### Installing RStudio

RStudio can be downloaded from the [posit.co website](https://posit.co/download/rstudio-desktop), which is the developer of RStudio. Make sure to pick the latest version available for your operating system.

Once you have [installed R and RStudio](04-working-with-r.qmd), you can start by launching RStudio. If everything was installed correctly, RStudio will automatically launch R as well.

The first time you open RStudio, you will likely see three separate windows. The first thing you want to do is **open an R Script (!!)** to work in. To do so, go to the toolbar and select **File -\> New File -\> R Script**.

You will now see four windows split evenly over the four corners of your screen:

-   In the **top-left** you have the text editor for the file that you are working in. This will most of the time be an R script or RMarkdown file.
-   In the **top-right** you can see the data and values that you are currently working with (environment) or view your history of input.
-   In the **bottom-left** you have the console, which is where you can enter and run code, and view the output. If you run code from your R script, it will also be executed in this console.
-   In the **bottom-right** you can browse through files on your computer, view help for functions, or view visualizations.

While you can directly enter code into your console (bottom-left), you should always work with R scripts (top-left). This allows you to keep track of what you are doing and save every step.

## Using RStudio

### Running code

Copy and paste the following example code into your R Script. For now, don't bother understanding the syntax itself. Just focus on running it.

```{r running-code-from-the-r-script-1, eval=F}
3 + 3
2 * 5
6 / 2
"some text"
"some more text"
sum(1,2,3,4,5)
```

You can **run** code by selecting the code and clicking on the **Run** button in the toolbar. However, we highly recommend getting used to using the keyboard shortcut, because this will greatly speed up your process. On Windows (and Linux) the shortcut is **Ctrl + Enter**. On Mac it's **Command + Enter**.

There are two ways to run code:

-   If you select a specific piece of code (so that it is highlighted) you can run this specific selection. For example, select the first three lines (the three mathematical operations) and press Ctrl + Enter. This should then print the results for these three mathematical expressions. Note that you can also select a specific part on a line. Try selecting just the second `3` on the first line. This should just print the number 3.
-   If you haven't made a selection, but your text cursor is somewhere on a line in your editor, you can press Ctrl + Enter to run the line where the cursor is at. This will also move the cursor to the next line, so you can *walk* through the code from top to bottom, running each line. Try starting on the first line, and pressing Ctrl + Enter six times, to run each line separately.


### Using RStudio projects

It is best to put all your code in an RStudio *project*. This is essentially a folder on your computer in which you can store the R files and data for a project that you are working on. While you do not necessarily need a project to work with R, they are very convenient, and we strongly recommend using them.

To create a new project, go to the top-right corner of your RStudio window. Look for the button labeled **Project: (None)**. Click on this button, and select New Project. Follow the instructions to create a new directory with a new project. Name the project "R introduction".

Now, open a new R script and immediately save it (select File -\> Save in the toolbar, or press ctrl-s). Name the file **my_first_r_script.r**. In the bottom-right corner, under the **Files** tab, you'll now see the file added to the project. The extension **.r** indicates that the file is an R script.


## Names and objects

### The power of assignment

In R, and in computer programming in general, the most essential operation is to assign **objects** to **names**.
By **object**, we then broadly mean any piece of information. a single number, a text, a list of numbers, and even an entire data set.

In plain terms, **assignment** is how you make R remember things by assigning them to a name. To assign an object to a name, we use the **arrow notation**: `name <- value`. For example:

```{r the-power-of-assignment-1}
x <- 2
```

::: {.callout-note title="Good to know: you can also use `x = 2`" collapse="true"}
Instead of using the **arrow** notation, you can also use the **equal sign** notation: `name = object`.

```{r the-power-of-assignment-2}
x = 2
```

We will in general always use the **arrow** notation. But if you encounter the **equal sign** notation, just remember that it's the same thing.
:::

By **running** the code `x <- 2`, you are saying: **Assign the value `2` to the name `x`**.
Any objects that you assigned to names are stored in your **Environment**. You can see this environment in the top-right window, under the **Environment** tab.
If you assigned `2` to `x`, you should see a table with in the left column the names (`x`) and in the right column a description of the object.
For simply objects like numbers, this will just be the value (`2`).

From hereon, when you use the name `x` in your code, it will **refer** to the value `2`. So when we run the code `x * 5` (x times 5) it will print the number `10`

```{r the-power-of-assignment-3}
x * 5
```

::: {.callout-note title="Why does R print a [1] before the result?" collapse="true"}
When running `x * 5`, R correctly prints the value 10. But why does it say `[1] 10`?
The reason is that R always thinks of a number (or string) as a **vector** (i.e. list of values), that can have 1 or multiple values.
The `[1]` indicates that 10 is the first (and only) value.

If you print a longer **vector**, you can see that R prints `[...]` at the start of each line, just to help you see the position of individual values.
The following code generates a vector with numbers from 1 to 50

```{r the-power-of-assignment-4}
1:50
```
:::



#### Assigning versus printing

Notice that when you ran the code `x <- 2`, R didn't print any values to the console (the bottom-left window).
But when you ran `x * 5`, R did print the value `10`.
Basically, when you run code, and you DO NOT assign the result to a name, R will print the result to the console.

So the following code will NOT print the object (`"I will not be printed"`) to the console, but will store it (you can see the name pop up in the **Environment** tab)

```{r assigning-versus-printing-1}
i_will_be_remembered <- "I will not be printed"
```

And the following object (`"I will be printed"`) will be printed to the console, but not stored in the **Environment**.

```{r assigning-versus-printing-2}
"I will be printed"
```



#### Assigning different types of objects

You can assign any type of object to a name, and you can use any name, as long as it starts with a letter and doesn't contain spaces or symbols (but underscores are OK)

```{r assigning-different-types-of-objects-1}
a_number = 5
my_cats_name = "Hobbes"
```

If you run this code and check you **Environment** (top-right), you should now see these name-object pairs added.


#### Assigning results

Till now we only directly assigned objects to names. This is convenient, but the power of assignment really shines when you use it to store **results**.
For example, we can also do this.

```{r assigning-results-1}
x = 5 + 10
```

This a very simple example, but just think for a second what this allows us to do.
Since we can assign anything to a name, we can break down any complicated procedure into multiple steps!
For now, the key lesson is just to wrap your head around the syntax for assigning objects to names.
This is fundamental to everything you will be doing in R (and in programming in general).

## Types of objects

### Common types of objects

You can work with many types of data in R.
Here are some of the most common types of objects you'll encounter:

- **Numeric**: Numbers, like `5`, `3.14`, or `-0.5`.
- **Character**: Text, like `"Hello, world!"`.
- **Factor**: Categorical data, like `education_level` or `country`.

Let's see how these types of objects work in practice.

```{r common-types-of-objects-in-r-1, results='hide'}
number <- 5
number

character <- "Hello, world!"
character

factor <- factor(c(1,2,2,3), labels = c("A", "B", "C"))
factor
```

#### Types determine what you can do with an object

The type of an object determines what you can do with it.
For example, you can perform mathematical operations on numeric objects, but not on character objects.

```{r types-determine-what-you-can-do-with-an-object-1, eval=F}
10 + 10    # returns 20
"10" + 10  # throws an error
```

#### You can coerce objects to different types

Sometimes you have an object of the wrong type. For instance, your numeric data might have been read in as a character object.

```{r you-can-coerce-objects-to-different-types-1}
number <- "5"
```

If I want to perform mathematical operations on `number`, I need to first convert it to a numeric object.
You can do this using the `as.numeric` function.

```{r you-can-coerce-objects-to-different-types-2, results='hide'}
number <- as.numeric(number)
number
class(number)    # numeric
```

Note that you cannot always convert objects to a different type.
Just use your common sense here.

```{r you-can-coerce-objects-to-different-types-3, results='hide', warning=FALSE}
as.numeric("I am not a number")
```

When coercion is not possible, R will return `NA` (missing value) and give you a warning.


### Vectors and Data Frames

In practice, you'll rarely work with single values alone.
However, many operations you can perform on a single value can also be applied to multiple values at once.
Understanding how individual values combine to form larger data structures is key to working effectively with data.

#### From Single Values to Vectors

A vector is a collection of values of the same type (e.g., all numeric or all character). You create a vector by combining individual values using the `c()` (combine) function.
Just like with single numbers, a vector has a type.

```{r from-single-values-to-vectors-1}
numbers <- c(1, 2, 3, 4, 5)
class(numbers)
```

Now, you can perform operations on the entire vector at once:

```{r from-single-values-to-vectors-2}
numbers * 10
sum(numbers)
```




#### From Vectors to Data Frames

Single vectors are still not very usefull for data analysis, since we're often interested in relations between variables.
This is where data frames come in.

A data frame is a table with rows and columns, where each row is an observation and each column is a variable.
You can think of a data frame as a collection of vectors, where each vector represents a column in the dataset.
We can create a data frame by combining vectors using the `data.frame()` function.[^1]

[^1]: Throughout this book we will actually be using the `tibble` data structure from the `tidyverse` package, which is an improved version of the `data.frame`.

```{r from-vectors-to-data-frames-1}
country <- c("NL", "NL", "BE", "BE", "DE", "DE", "FR", "FR", "UK", "UK")
height  <- c(176 , 165 , 172 , 160 , 180 , 170 , 175 , 165 , 185 , 175 )

d <- data.frame(country, height)
```

Now, you can perform operations on the entire data frame at once!
You can for instance perform a statistical test to see if the average height of people differs between countries.
Just like with single values and vectors, we need to take the type of each column into account.
For instance, we can't perform a correlation analysis using the `country` column, because it's a character vector.

## Functions

### What is a function? {#what-is-a-function}
99% of what you do in R will involve using functions.
A function in R is like a mini-program that you can use to perform specific tasks.
It takes *input*, processes it, and gives you an *output*.
For example, there are functions for:

* importing data
* computing descriptive statistics
* performing statistical tests
* visualizing data

A function in R has the form:

```{r what-is-a-function-1, eval=F}
output <- function_name(argument1, argument2, ...)`
```

* **function_name** is a name to indicate which function you want to use. It is followed by parentheses.
* **arguments** are the input of the function, and are inserted within the parentheses.
* **output** is anything that is returned by the function.

For example, the function `c` *c*ombines multiple values into a *vector*.

```{r what-is-a-function-2}
x = c(1,2,3,4)
```

Now, we can use the `mean` function to calculate the mean of these numbers:

```{r what-is-a-function-3}
m <- mean(x)
```

The calculated mean, `2.5`, is now assigned to the name `m`:

```{r what-is-a-function-4}
m
```

#### Optional arguments {#optional-arguments}

In the `c` and `mean` functions above, all the arguments were **required**.
To **c**ombine numbers into a vector, we *needed* to provide a list of numbers.
To calculate a **mean**, we *needed* to provide a numeric vector.

In addition to the **required** arguments, a function can also have **optional** arguments, that give you more control over what a function does.
For example, suppose we have a range of numbers that also contains a missing value. In R a missing value is called `NA`, which stands for `Not Available`:

```{r optional-arguments-1}
x_with_missing <- c(1, 2, 3, NA, 4)
```

Now, if we call the `mean` function, R will say that the mean is unknown, since the third value is unknown:

```{r optional-arguments-2}
mean(x_with_missing)
```

This is statistically a very correct answer.
But often, if some values happen to be missing in your data, you want to be able to calculate the mean just for the numbers that are not missing.
Fortunately, the mean function has an **optional argument** `na.rm` (remove NAs) that you can set to `TRUE` (or to `T`, which is short for `TRUE`) to ignore the NAs:

```{r optional-arguments-3}
mean(x, na.rm=TRUE)
```

Notice that for the **required** argument, we directly provide the input `x`, but for the **optional** argument we include the argument name `na.rm = TRUE`.
The reason is simply that there are other optional arguments, so we need to specify which one we're using.

::: {.callout-note title="How do I know what arguments a function has?" collapse="true"}

To learn more about what a function does and what arguments it has, you can look it up in the 'Help' pane in the bottom right,
or run `?function_name` in R.

```{r optional-arguments-4, eval=F}
?mean
```

Here you can learn about the `na.rm` argument that we just used!

If you are just getting to know R, we recommend first finishing the rest of the `Getting Started` section.
Then once you get the hang of things, have a look at the [Use ?function help page](04-working-with-r.qmd) tutorial.
:::


#### Using the pipe syntax {#pipe-syntax}

There is another common way to use functions in R using the **pipe** syntax.
With the pipe syntax, you can *pipe* the first argument into the function, instead of putting it inside the parentheses.
As you will see below, this allows you to create a pipeline of functions, which is often easier to read.
<!-- The pipe syntax uses the `|>` operator, which is used as follows: -->

```{r using-the-pipe-syntax-1, eval=F}
argument1 |> function_name(argument2, ...)
```

For example, the following two lines of code give identical output:

```{r using-the-pipe-syntax-2}
mean(x_with_missing, na.rm=T)

x_with_missing |> mean(na.rm=T)
```

Notice how our first argument, the **required** argument `x_with_missing`, is *piped* into the mean function.
Inside the mean function, we only specify the second argument, the **optional** argument `na.rm`.

So why do we need this alternative way of doing the same thing?
The reason is that when writing code, you shouldn't just think about **what the code does**, but also about **how easy the code is to read**.
This not only helps you prevent mistakes, but also makes your analysis **transparent**.
As you'll see later, you'll encounter many cases where your analysis requires you to string together multiple functions.
In these cases, pipes make your code much easier to read.

For example, imagine we would want to round the result (`2.5`) up to a round number (`3`).
With the pipe syntax we can just add the round function to our pipeline.

```{r using-the-pipe-syntax-3, results=F}
x_with_missing |>
  mean(na.rm=T) |>
  round()
```

You'll see how powerful this can be later on, especially in the [Data Management](07-data-frames.qmd) chapter.
In order to prepare and clean up your data, you'll often need to perform a series of functions in a specific order.
The pipe syntax allows you to do this in a very readable way.

### Mastering functions

There are some usefull tricks for using functions in R that are good to know about.
We do not discuss these here, because if you're just starting out, there are more important things to learn first.
But once you get the hang of things, you can learn more about these tricks in the [Good to Know](04-working-with-r.qmd) tutorial.

## Packages

### What is a package? {#what-is-a-package}


In R, a package is a collection of functions, data, and documentation that extends the capabilities of R.
You can think of packages kind of like apps on your phone: they provide additional functionality that you can use to perform specific tasks.
Also, like apps, you can install and uninstall packages as needed, directly from within R.

There are thousands of R packages available, which enables you to use almost any existing data analysis technique.
R is not just a tool for statistical analysis, but also for data visualization, data collection, articifial intelligence, and much more.
If there is anything you need to do, there is a good chance that someone has already written a package for it.

### How to install

To use a package, you first need to install it. **Installing a package only needs to be done once** per system (unless you need to update it).

#### 1. Install a package

Most packages are available on the Comprehensive R Archive Network (CRAN), which is the main repository for R packages.
To install these packages, all you need to know is their name.

For example, there is a package called `lubridate` that makes it easier to work with dates and times in your data.
To install this package, you can use the `install.packages()` function:

```{r 1-install-a-package-1, eval=F}
install.packages("lubridate")
```

::: {.callout-warning title="What if R asks tot 'install from source'?" collapse="true"}

When running `install.packages()` You sometimes get the message that `There is a binary version available but the source version is later` (we're mainly seen this on Mac). You then get the question whether you want to `install from source` the package which needs compilation `(Yes/no)`.
To answer this question, you have to type "yes" or "no" in your R console.

It is then usually **best to say NO**.
This will install a slightly older version of the package, but it will be much faster and easier to install.
You often don't need the latest version, so it's not worth the extra hassle.

In case you're curious, the reason for this is that the newest version has not been *prepared* for your system yet.
They do have the **source code**, but it has not yet been compiled into a **binary** version that is ready to use.
Think of the source code as a recipe, and the binary version as the ready-made dish.
If you really want to have the newest version you can say "yes", but you'll have to cook it yourself!
The main problem is that you will often need to install some extra software to do this, which can be a hassle.
So unless you really need the newest version, it's usually best to say "no", and just install the older version that has already been prepared for your system.
:::

#### 2. Load a package

Once you have installed a package, it is not yet loaded into your current R session.
Similar to when you install a new app on your phone, you need to open it every time you want to use it.

To use a package in your current R session, you can load it with the `library()` function:

```{r 2-load-a-package-1, message=FALSE}
library(lubridate)
```

### Package documentation

Most packages come with good documentation that explains how to use the functions in the package.
For individual functions you can use the `?` operator to open the help page for that function, as we explain in [here](04-working-with-r.qmd).
But there is often also a more general documentation that explains the package as a whole.
This is called a **vignette**, and if a package has one, you can open it with the `vignette()` function.

```{r how-do-i-know-how-to-use-a-package-1, eval=F}
vignette("lubridate")
```

You can look up vignettes and function documentation [online](https://www.rdocumentation.org/).
This is also a good way to find new packages for whatever you're trying to do.
