---
title: "1.4 Base R Fundamentals"
subtitle: "An optional reference guide to vectors, indexing brackets, and custom functions"
---

# Why Learn Base R?

In this course, we prioritize the **tidyverse** (`dplyr`, `ggplot2`) because of its intuitive, expressive syntax for data manipulation.
However, R is fundamentally built on **Base R**: the built-in language structures that have powered R for decades.

You will encounter Base R when reading statistical package documentation, viewing R help pages, or inspecting code on StackOverflow.
This optional reference covers the core Base R conventions every student should recognize.

::: {.callout-note title="Optional Material"}
This module is optional for the core course curriculum. You can complete the mandatory assignments using only tidyverse syntax.
:::

# 1. Vectors and Data Types

In Base R, the atomic building block is the **vector** created with `c()`:

```{r}
numbers <- c(10, 20, 30, 40, 50)
fruits <- c("apple", "banana", "cherry")
flags <- c(TRUE, FALSE, TRUE)
```

Common types:
- `numeric` (integers and floating-point doubles)
- `character` (text strings)
- `logical` (`TRUE` or `FALSE`)
- `factor` (categorical variables with predefined levels)

# 2. Square Bracket Indexing `[ ]`

While dplyr uses `filter()` and `select()`, Base R subsets vectors and matrices using square brackets:

```{r}
# 2nd element
numbers[2]

# Elements 1 through 3
numbers[1:3]

# Logical filtering
numbers[numbers > 25]
```

## Data Frame Indexing: `[row, column]`

For 2D tables, Base R uses `[rows, columns]`:

```{r}
df <- data.frame(
  id = 1:3,
  score = c(85, 92, 78),
  grade = c("B", "A", "C")
)

# Row 1, Column 2
df[1, 2]

# All rows where score > 80
df[df$score > 80, ]
```

# 3. Dollar Sign Operator `$`

The `$` operator extracts a single column as a vector:

```{r}
df$score
mean(df$score)
```

# 4. Writing Custom Functions

You can define reusable functions in Base R using the `function()` keyword:

```{r}
calculate_zscore <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

calculate_zscore(df$score)
```
