---
title: "Factor analysis"
---

## Why Factor Analysis?

Many concepts in the social sciences are too complex to be captured with a single question. Think of *academic motivation*, *media literacy*, or even *happiness*. Each of these is a **latent construct**: something that cannot be measured directly, but only through multiple related items. In surveys you won’t encounter a direct statement like: "My academic motivation is high" with a Likert scale. Instead, you might see a range of related statements such as "I look forward to studying," "I keep working even when assignments are difficult," or "I set goals for my learning." Taken together, these items give us a better picture of the underlying latent construct. But what if these items reflect not one, but *multiple* underlying constructs?

Exploratory Factor Analysis (EFA) is a statistical technique that helps us figure out **how many underlying dimensions** (factors) exist in our data and **which items belong together**. In other words, it helps us see whether our items really measure one construct, or whether multiple related constructs are at play.

Imagine a researcher wants to capture respondents’ media literacy. Instead of asking one direct question like “How would you rate your media literacy?”, they design a series of concrete items. In this chapter, we’ll work with those items. Respondents rated the following ten statements on a 1–5 scale, ranging from 'Strongly Disagree' to 'Strongly Agree':

1.  I can distinguish between facts and opinions in the news.
2.  I share or post political content on social media.
3.  I recognize credible and non-credible news sources.
4.  I notice when news stories present only one side of an issue.
5.  I look at the date of an article before judging its relevance.
6.  I actively seek out diverse political viewpoints.
7.  I critically reflect on the framing of political issues.
8.  I check who published a news article before trusting it.
9.  I consider the intention behind a media message.
10. I discuss political content I see online with others.

Take a moment to read through these ten statements yourself: do you spot any hidden patterns? Do you believe that they all reflect one construct of media literacy, or can you see them group into multiple factors (possibly different *kinds* of media literacy), based on the topics they tap into? This is exactly what EFA does, but based on math, rather then substantive interpretation: it looks at the patterns of correlations between items and groups them into meaningful clusters, that you will be able to interpret!

[You might have heard about or later on encounter*Principal Component Analysis (PCA)*. While PCA also reduces data into fewer dimensions, it has a different goal (data reduction rather than uncovering latent constructs). In this course, we focus only on **factor analysis**.]{style="font-size:70%"}

### Principal Axis Factoring

There are various methods and approaches in factor analysis. In this chapter we use the method of **Principal Axis Factoring (PAF)**. The underlying model assumes that the correlations we observe between items are not direct causal effects between those items, but are explained by one or more hidden common factors. A metaphor: think of the items as puppets on a stage that seem to move together. PAF looks for the invisible strings above the stage. Once you remove those hidden strings, the puppets no longer move in sync. In statistical terms: if you remove the effect of the factor(s), the correlations between items disappear as much as possible.

### Factor analysis for scale construction

Before we dive into the practical analysis, it’s helpful to place factor analysis within the broader process of scale construction. When building a scale, researchers typically: (1) identify indicators, (2) check the direction and distributions of the items, (3) determine the dimensionality with factor analysis, (4) test internal consistency with reliability analysis, (5) compute the scale(s) and (6) validate the scale(s). The main focus of this chapter is is Step 3: using factor analysis to uncover the dimensionality of our items.

In this chapter, we will work with the *Media Literacy* scale from our simulated *Media & Democracy* dataset (available on Canvas as `polcomm_data.csv`). Ten items (`ml_1` to `ml_10`) were designed to measure media literacy.

Our key question is: *Do these ten items measure a single underlying construct (general media literacy), or are there multiple dimensions hidden within them?*

The answer to that question will then guide us further in to process of scale construction. Let's find out how we can best measure media literacy! 

### Step 1: Inspect your data

We load the dataset `polcomm_data.csv` - which is a simulated dataset (N = 200) containing various variables on media use, democracy and related topics - and name it `d`.

Before any analysis, it’s good practice to run descriptive statistics. This shows whether the items are coded consistently, whether all answer categories were used, and if there are missing values.

```{r step-1-inspect-your-data-1, message=FALSE}
library(tidyverse) 
library(summarytools)

d <- read_csv("https://data.dataliteracy.cc/simulated/polcomm_data.csv")

### Inspect Media Literacy items
d |> select(ml_1:ml_10) |> descr()

```

The output shows the means, standard deviations, and ranges. In our case, all items are measured on a 1–5 Likert scale, with no missing values. The averages hover mostly around 3–4, which makes sense for survey data.

## Step 2: Check correlations

Factor analysis is based on **correlations** between items. If items do not correlate, they probably don’t measure the same construct. And if an item shares no correlations at all, it has nothing to do with the other items, so does not fit in the measure of the construct. On the other hand, if correlations are *too high* (above 0.8), it could indicate redundancy.

```{r step-2-check-correlations-1, message = F}
library(sjPlot)

d |>
  select(ml_1:ml_10) |>
  tab_corr(p.numeric = TRUE)
```

Here we see a healthy mix: many correlations fall between 0.3 and 0.7, which is a sign that these items are related but not identical. At the same time, the output also shows a few very low correlations (close to 0), meaning that some items barely share variance with each other. This is not necessarily a problem, as long as each item correlates with at least some others. In fact: it may point to different dimensions of the concept, which is why we perform factor analysis after all!

We also spot one or two correlations that are quite high (above 0.8), which could indicate that certain items are almost redundant — they may be tapping into exactly the same skill. Together, this pattern suggests that the items can form meaningful clusters, while also raising questions about whether some pairs overlap too much or too little.

## Step 3: How many factors?

Before the actual factor analysis, that will give us information about the content of our factors, we would like to know how many factors we should extract in the first place.

For this, we use two common criteria:

-   **Scree plot / Cattell criterion**: look at a plot of eigenvalues in descending order. The blue line will usually flatten out at some point. The "elbow" or kink indicates where meaningful factors stop and noise begins. Following the rule of thumb, you take the kink point minus one as the number of factors to extract.
-   **Kaiser criterion**: check the eigenvalues themselves. Each eigenvalue reflects how much variance a factor explains. According to Kaiser, only factors with an eigenvalue greater than 1 (meaning they explain at least as much variance as a single item) should be kept. Factors below 1 are considered too weak to retain. In the output we look at the `eigen values of components`.

```{r step-3-how-many-factors-1, message=F}
library(psych)

d |>
  select(ml_1:ml_10) |>
  fa.parallel(fa = "fa", fm = "pa") |> print()
```

Both the scree plot and the eigenvalues suggest that a **three-factor solution** is appropriate. If the criteria yield two different outcomes, we prioritize the scree plot. At the same time, remember there is also a third - less formal - criterion: **interpretability**. The mathematical rules give us a suggestion for the number of factors, but as researchers we need to check whether the resulting factors make substantive sense. If we cannot meaningfully interpret the clusters of items, the factor solution is not useful, even if the numbers suggest it is. This highlights that factor analysis is both a statistical and a substantive exercise.

## Step 4: Run the factor analysis

Now we can actually run the factor analysis. We specify:

-   **`nfactors = 3`** → we want R to extract three factors

-   **`rotate = "oblimin"`** → an oblique rotation, allowing factors to correlate (realistic in social sciences)

    #### **Rotation**

    For reasons of interpretability (i.e., to appoint substantive meaning to the factors), factors are (often) **rotated**. Visualize a multi-dimensional space, in which each factor represents an axis (e.g. X or Y-axis). Each item is located somewhere in this space, based on their loadings on the respective factors. Rotation changes how the factors are aligned, which makes the loadings easier to interpret. Without rotation, items might load moderately on multiple factors, making it hard to see clear clusters. By rotating, we maximize loadings on one factor while minimizing them on others.

    Broadly, there are two approaches to this: **orthogonal rotation** and **oblique rotation**. Orthogonal rotation (e.g. varimax) keeps the factors uncorrelated, with axes at a 90° angle. Oblique rotation (e.g. oblimin), which we use here, allows factors to correlate with each other. In social sciences, this is usually the more realistic choice, because underlying constructs often overlap to some extent.

-   **`fm = "pa"`** → principal axis factoring (PAF), the method of factor analysis that is very common in EFA.

```{r step-4-run-the-factor-analysis-1, message = F}
fa_result <- d |>
  select(ml_1:ml_10) |>
  fa(nfactors = 3, rotate = "oblimin", fm = "pa")

print(fa_result)
```

The output includes a **pattern matrix** (the first table you encounter titled `Standardized loadings`), which shows how strongly each item loads onto each factor. In Principal Axis Factoring, an item ‘loading’ strongly on a factor means that the variance in that item can be largely explained by that underlying factor. The loading is essentially a correlation: a high positive loading indicates that when the factor increases, the likelihood of agreeing with that item also increases. In other words, the factor is the hidden thread, and the loading shows how tightly the item is tied to it. increases,

Note, that when factors are *correlated* (when using oblimin rotation), loadings represents the relationship of an item with the factor, *controlled for* the other factors.

By default, the pattern matrix includes even very small loadings. However, interpretation becomes easier when we would only show the higher factor loadings. We usually keep only loadings above 0.3. This makes it easier to see clusters of items:

```{r step-4-run-the-factor-analysis-2}
fa_result$loadings |> print(cutoff = 0.3)
```

Now, the clusters are clear:

-   Factor 1: `ml_1`, `ml_4`, `ml_7`
-   Factor 2: `ml_3`, `ml_5`, `ml_8`, `ml_9`
-   Factor 3: `ml_2`, `ml_6`, `ml_10`

We can also visualize this:

```{r step-4-run-the-factor-analysis-3}
fa.diagram(fa_result)
```

This diagram shows factors as circles, items as boxes, and arrows representing strong (\> .3) loadings. The arrows indicate which items are strongly linked to which factors, based on their loadings in the pattern matrix. This visual is a direct translation of the principal axis factoring model you learned about earlier: the circles represent the hidden factors (the "invisible strings"), and the arrows represent how strongly those factors pull on each observed item. Because we applied an oblimin rotation, some circles may be connected with arrows, reflecting that the factors themselves can correlate. In short, the diagram helps you see at a glance which items cluster together under each latent construct and how those latent constructs may relate to each other.

## Step 5: Interpret the factors

It’s up to the researcher to give meaning to the factors. Take another look at the ten Media Literacy items and the clusters in the pattern matrix. Can you group them yourself? Which items seem to belong together, and what label would you give to each group? Try it out – this interpretive step is part of what makes factor analysis more than just math.

This is how you might have interpreted the factors:

-   *Factor 1* (`ml_1`, `ml_4`, `ml_7`): **Analytical media literacy** includes items about reflecting critically on content (distinguishing fact/opinion, spotting one-sidedness, reflecting on framing).
-   *Factor 2* (`ml_3`, `ml_5`, `ml_8`, `ml_9`): **Source literacy** includes items about judging source credibility (who published it, intention, date, credibility).
-   *Factor 3* (`ml_2`, `ml_6`, `ml_10`): **Active engagement** (as a part of media literacy) includes items about sharing, seeking diverse views, and discussing content online.

So, instead of one single Media Literacy scale, our analysis suggests there should be **three distinct but related dimensions** that each have a meaningful interpretation.

## Step 6: Reliability of each factor

Once we have identified factors, we need to check whether the items in each factor form a reliable scale, in other words: whether the scales are internally consistent. This is where **Cronbach’s alpha** comes in.

We won’t go into detail here, since we have a full chapter on [Scale construction](https://dataliteracy.cc/analysis/techniques/scale-construction.html), but here is how you would do it for the scale arising from Factor 1:

```{r step-6-reliability-of-each-factor-1}
### Example for the Analytical factor
d |>
  select(ml_1, ml_4, ml_7) |>
  psych::alpha()
```

The output will show that all three scales (Analytical, Source, Active) have high reliability (alpha \> .8).

## Step 7: Construct the scales

Finally, we compute new scale scores by averaging the items in each factor. In practice, this means creating new variables (`ml_reflect`, `ml_source`, `ml_active`) that take the mean of the items belonging to each factor. For example, the variable `ml_reflect` is calculated by summing items `ml_1`, `ml_4`, and `ml_7` and dividing by 3 to compute an average that stays on the same 1–5 scale. This way, each respondent’s answers on individual items are combined into a single, interpretable score that represents their position on that underlying dimension.

```{r step-7-construct-the-scales-1}
d <- d |>
  mutate(ml_reflect = (ml_1 + ml_4 + ml_7)/3,
         ml_source  = (ml_3 + ml_5 + ml_8 + ml_9)/4,
         ml_active  = (ml_2 + ml_6 + ml_10)/3)

### Inspect the new scales
d |> select(ml_reflect, ml_source, ml_active) |> descr()
```

We now have three new variables that represent distinct aspects of media literacy. These can be used in further analysis, depending on the theoretical framework and research question.

## Summary

Factor analysis is a powerful tool for uncovering the **dimensionality** of scales:

-   It helps us determine whether a set of items measures one construct or multiple constructs.
-   We decide on the number of factors using the scree plot and eigenvalues of components, but also check whether the solution is *substantively interpretable*.
-   We interpret factors by looking at clusters of items with strong loadings and assigning them meaningful labels.
-   Finally, we check for the internal consistency of our scales, before actually constructing them and applying them in an analysis.

In our Media Literacy example, instead of one general scale, we uncovered three meaningful dimensions: **Analytical reflection**, **Source literacy**, and **Active engagement**. This shows how factor analysis provides a richer and more accurate way of measuring how people understand and interact with media.
