---
title: "2.2 Group and summarize"
subtitle: "Calculating and interpreting summary statistics."
---

*You arrive here from [Data Frames](2.01-data-frames.qmd), so you can already load a data frame and pick the columns and rows you need. Here you'll group that data and summarise it per group with `group_by()` and `summarise()`. Comparing groups like this is at the heart of a lot of research, and it's also the stepping stone to descriptive statistics and plotting.*

→ [See this module in the skill tree](curriculum.qmd?module=summa)

# Data Summarization

The functions used in earlier parts on data preparation worked on individual rows.
Sometimes, you need to compute properties of groups of rows (cases).
This is called aggregation (or summarization) and in the tidyverse uses the `group_by` function
followed by either `summarize` or `mutate`.

We'll use a real public dataset for this: polling data on support for stricter gun laws, broken down by question and political party.

```{r, message=F, warning=F}
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
d <- read_csv(url) |>
  select(-URL) |>
  rename(Rep = `Republican Support`, Dem = `Democratic Support`)
d
```

# Grouping rows

We can use the `group_by` function to group data by a specific column, for example by `Question`:

```{r}
d |>
  group_by(Question)
```

The data itself hasn't changed its values, but dplyr records that the data is now grouped by `Question` (with 8 groups in total).

# Summarizing

To summarize grouped data, follow `group_by` with a call to `summarize`.
The syntax is `summarize(column = calculation, ...)`.
Crucially, the calculation must compute a single summary value given a vector of values.
Very common summarization functions are `mean()`, `sd()`, `median()`, and `n()`.

For example, the following computes the average support per question:

```{r}
d |>
  group_by(Question) |>
  summarize(Support = mean(Support, na.rm = TRUE)) |>
  arrange(desc(Support))
```

`summarize` reduces the rows equal to the number of groups (8), keeping only the grouping variables and the newly calculated summary columns.

You can also compute summaries of multiple values:

```{r}
d |>
  group_by(Question) |>
  summarize(
    Dem = mean(Dem, na.rm = TRUE),
    Rep = mean(Rep, na.rm = TRUE),
    diff = mean(Dem - Rep, na.rm = TRUE)
  ) |>
  arrange(desc(diff))
```

Another essential helper is `n()` (without arguments), which counts the number of rows in each group:

```{r}
d |>
  group_by(Question) |>
  summarize(
    n = n(),
    mean = mean(Support, na.rm = TRUE),
    sd = sd(Support, na.rm = TRUE)
  )
```

# Using mutate with group_by

While `summarize` collapses multiple rows into a single summary row, using `mutate` after `group_by` preserves all rows and adds the group-level summary as a new column:

```{r}
d2 <- d |>
  group_by(Question) |>
  mutate(
    avg_support = mean(Support, na.rm = TRUE),
    diff = Support - avg_support
  )
d2
```

# Ungrouping

Finally, you can use `ungroup()` to remove grouping metadata. When you are done computing grouped operations, calling `ungroup()` ensures subsequent functions treat the data frame as a single whole:

```{r}
d2 |>
  ungroup() |>
  summarize(overall_sd = sd(diff, na.rm = TRUE))
```
