---
title: "3.1 Means & Variance"
subtitle: "Understanding the mean and variance of a variable."
---

*You arrive here from [R language](1.02-r-language.qmd), so you can already compute things in R with vectors and functions. Here you'll learn what the mean, variance, and standard deviation of a variable actually tell you, and how to compute them, by hand once and then with `mean()`, `var()`, and `sd()`. This matters because nearly every statistic you'll meet later in the book — t-tests, correlations, regression coefficients — is built on top of these two numbers.*

→ [See this module in the skill tree](curriculum.qmd?module=me-va)

::: {.callout-warning title="AI-assisted draft"}
This content was drafted by an AI assistant and has not yet been manually checked by course staff. Treat it as a starting point, not a verified source.
:::

## The mean: a balance point

Suppose we asked 10 respondents to rate their political interest on a scale from 1 (not interested at all) to 10 (very interested):

```{r}
library(tidyverse)

political_interest <- c(2, 3, 3, 4, 5, 5, 6, 7, 8, 9)
```

The **mean** is the number you already know as "the average": add up all the values, and divide by how many there are.

$$ \bar{X} = \frac{\sum X_i}{n} $$

```{r}
sum(political_interest) / length(political_interest)
```

Of course, R has a built-in function for this, so you'll never actually need to type out the formula:

```{r}
mean(political_interest)
```

It helps to think of the mean as a **balance point**. If you drew the 10 scores as weights on a see-saw, the mean is the point where the see-saw would balance. Some respondents are far below the mean, some are far above it, and the positive and negative distances exactly cancel out.

This "distance from the mean" is called a **deviation**, and it's the building block for everything else in this chapter. For a given value $X_i$, the deviation is $X_i - \bar{X}$.

```{r}
d <- tibble(political_interest) |>
  mutate(deviation = political_interest - mean(political_interest))

d
```

If you sum the `deviation` column, you'll always get (approximately) zero — that's exactly what "balance point" means.

```{r}
sum(d$deviation)
```

## Variance and standard deviation: how spread out is the data?

The mean tells you where the *center* of the data is, but it tells you nothing about how spread out the scores are. Two samples can have the exact same mean while looking completely different: in one, everyone agrees and clusters close to the average; in the other, respondents are scattered from one extreme to the other. For a researcher, that difference matters a lot — a highly polarized sample is a different story than a uniformly moderate one.

To measure spread, we'd like to average the deviations from the mean. But as we just saw, the deviations always sum to (about) zero, so a plain average of them is useless — positive and negative deviations cancel out. The fix is to **square** each deviation first. Squaring makes every value positive, and it also makes large deviations count disproportionately more than small ones, which is a reasonable way to penalize scores that are far from the average.

That gives us the **variance**: the (roughly) average squared deviation from the mean.

$$ var(X) = \frac{\sum (X_i - \bar{X})^2}{n - 1} $$

```{r}
d <- d |> mutate(squared_deviation = deviation^2)
d

sum(d$squared_deviation) / (length(political_interest) - 1)
```

Again, R has a built-in function so you don't need to do this by hand in practice:

```{r}
var(political_interest)
```

### Why divide by $n - 1$ and not $n$?

You might expect we'd divide by $n$, the number of observations, since variance is supposed to be an *average* squared deviation. The reason we divide by $n - 1$ instead is a subtle but important one: we're almost always using a **sample** to estimate the variance of a **population**, and the sample mean $\bar{X}$ is, by construction, the value that makes the squared deviations in *this particular sample* as small as possible. That means a sample will systematically underestimate the true population variance just a little bit, on average. Dividing by the slightly smaller number $n - 1$ instead of $n$ corrects for this bias.

You don't need to be able to re-derive this to use it correctly — just remember the rule of thumb: when you're working with a sample (which is almost always the case in practice), use $n - 1$. R's `var()` and `sd()` functions already do this for you.

### Standard deviation: back to the original units

One awkward thing about variance is that it's expressed in **squared units**. If political interest is measured on a 1–10 scale, the variance is in "scale points squared", which is hard to interpret intuitively. The **standard deviation** fixes this by taking the square root of the variance, bringing it back to the original scale:

$$ SD(X) = \sqrt{var(X)} $$

```{r}
sqrt(var(political_interest))
sd(political_interest)
```

A useful (rough) intuition for the standard deviation: for many variables, most observations fall within about one SD of the mean, and nearly all fall within two or three SDs. So if political interest has a mean of 5.2 and an SD of 2.2, a respondent scoring 9 is unusually interested compared to the rest of the sample, while someone scoring 6 is close to typical.

## Why this matters for everything that follows

The mean and the variance (or SD) aren't just descriptive trivia — they are the two ingredients that almost every statistical test in this book combines in some way:

* A **t-test** (in [Comparing two Groups](5.01-comparing-two-groups.qmd)) compares the *difference between two means*, relative to how much variance there is within each group.
* A **correlation** (in [Correlation & Covariance](3.03-correlation-and-covariance.qmd)) is built from deviations from the mean of two variables at once.
* A **regression model** (in [Multiple Regression](5.06-regression.qmd)) tries to explain variance in your outcome variable using one or more predictors.
* **Effect sizes** (in [Effects](3.04-effects.qmd)), like Cohen's $d$, are literally mean differences expressed in standard deviation units.

In other words: once you understand the mean as "where the center is" and the variance/SD as "how much do individual scores typically deviate from that center", you already understand the core building block of most of the statistics you'll see from here on.

## Further resources {.unnumbered}

- 🎥 [StatQuest: Calculating the Mean, Variance and Standard Deviation, Clearly Explained!!!](https://www.youtube.com/watch?v=SzZ6GpcfoQY) — a short, visual walkthrough of exactly the calculations in this chapter.
