---
title: "4.3 Advanced Visualization"
subtitle: "Advanced visualization techniques: pooling, faceting, and interactive plots."
---

*You arrive here from [Data Visualization](4.02-data-visualization.qmd) and [Pivoting and joining](2.04-pivoting-and-joining.qmd), so you can already make a basic plot and reshape data into long format. Here you'll learn to split one plot into several small, comparable panels with faceting, combine multiple geoms in a single plot, and make deliberate choices about color and clarity so your plots are easy to read for everyone — including colorblind readers. These are the techniques that turn a quick exploratory plot into one you'd actually put in a report or a paper.*

→ [See this module in the skill tree](curriculum.qmd?module=vis-a)

::: {.callout-warning title="AI-assisted draft"}
This content was drafted by an AI assistant and has not yet been manually checked by course staff. Treat it as a starting point, not a verified source.
:::

## Faceting: small multiples

We'll continue with the same practice dataset:

```{r setup, message=FALSE}
library(tidyverse)
d <- read_csv("https://dev.dataliteracy.cc/practice_data.csv")
```

**Faceting** splits one plot into a grid of small panels, one per level of a categorical variable — so instead of squeezing several groups into a single crowded plot (e.g. by color), you compare them side by side. Add `facet_wrap(~ variable)` to any existing `ggplot2` call:

```{r facet-1}
ggplot(d, aes(x = trust_t1)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  facet_wrap(~experiment_group)
```

This is often clearer than overlapping histograms or densities, especially once you have more than two or three groups. You can facet by two variables at once with `facet_grid(row_var ~ col_var)`, which arranges panels in a full grid instead of wrapping them into rows.

Faceting also works well with **long-format data** (see [Pivoting and joining](2.04-pivoting-and-joining.qmd)). For example, say we want to compare trust at two time points (`trust_t1` and `trust_t2`) in one plot. First we pivot the two columns into a long format with a `time` and a `trust` column, then facet by `time`:

```{r facet-2, message=FALSE}
d_long <- d |>
  pivot_longer(cols = c(trust_t1, trust_t2), names_to = "time", values_to = "trust")

ggplot(d_long, aes(x = trust)) +
  geom_histogram(binwidth = 1, fill = "darkorange", color = "white") +
  facet_wrap(~time)
```

## Combining geoms

A single plot can layer multiple geoms on top of each other, as long as they share (or can use) the same aesthetic mapping. A common combination is points plus a fitted trend line, using `geom_smooth()`:

```{r combine-1, message=FALSE}
ggplot(d, aes(x = age, y = trust_t1)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "darkred")
```

Each geom is added as its own layer, drawn in the order you add it — here the points are drawn first, and the trend line on top. You can also combine a summary geom with the raw data, for example showing each group's mean trust as a point on top of individual observations:

```{r combine-2, message=FALSE}
group_means <- d |>
  group_by(experiment_group) |>
  summarise(mean_trust = mean(trust_t1, na.rm = TRUE))

ggplot(d, aes(x = experiment_group, y = trust_t1)) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  geom_point(data = group_means, aes(y = mean_trust), color = "red", size = 4)
```

Notice that the second `geom_point()` uses its own `data` argument (`group_means`) instead of the data set up in `ggplot()` — this is how you combine information computed at different levels (individual observations and group summaries) in one plot.

## Accessibility and clarity

A plot that looks fine to you may be unreadable to someone else. A few habits make a real difference:

* **Don't rely on color alone.** Around 1 in 12 men have some form of red-green color blindness. If color carries meaning, also vary the shape, line type, or facet so the distinction survives if color is lost (e.g. printed in grayscale).
* **Use colorblind-safe palettes.** The built-in `scale_color_viridis_d()` / `scale_fill_viridis_d()` (discrete) and `scale_color_viridis_c()` (continuous) are perceptually uniform and colorblind-friendly by default:

```{r accessibility-1}
ggplot(d, aes(x = experiment_group, y = trust_t1, fill = experiment_group)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal()
```

* **Avoid chartjunk.** Skip 3D effects, unnecessary gridlines, or decorative images — they add visual noise without adding information. A clean theme like `theme_minimal()` already removes most of this by default.
* **Label directly where you can.** For a small number of categories, a direct label next to each line or bar is often easier to scan than a separate legend the reader has to match back to the plot by color.

Interactive plots (for example with the `plotly` package, via `ggplotly()`) can add value for exploration — letting readers hover over points for exact values — but they don't work in a printed PDF or a static report, so use them sparingly and only for formats that support them (e.g. an html report or a dashboard).

## Further resources {.unnumbered}

- 📘 [ggplot2 reference: facet_wrap()](https://ggplot2.tidyverse.org/reference/facet_wrap.html) — the full set of options for faceting, including free scales and custom layouts.
- 📘 [viridis color scales](https://ggplot2.tidyverse.org/reference/scale_viridis.html) — documentation for the colorblind-friendly palettes used above.
