---
title: "5.2 Testing Associations"
subtitle: "Testing correlation coefficients for significance."
---

*You arrive here from [Hypothesis Testing](3.05-hypothesis-testing.qmd), so you already know how hypothesis testing works, and you can compute a correlation to describe a relationship. Here you'll test whether a correlation in your sample is statistically different from zero with `cor.test()`. A correlation in a sample can arise by chance, so testing it tells you how much confidence you can place in the relationship — and it prepares you for the linear model, since `cor.test()` turns out to be a t-test in disguise.*

→ [See this module in the skill tree](curriculum.qmd?module=te-as)

## Example data

We use the same small toy data set as in [Correlation & Covariance](3.03-correlation-and-covariance.qmd):

```{r}
library(tidyverse)
d <- tibble(x = c(20, 10, 30, 40, 50),
            y = c( 3,  2,  5,  4,  6))

d
```

## Testing a correlation with `cor.test()` {#testing-correlation}

In empirical research, we do not merely calculate the sample correlation $r$; we also test whether the correlation is statistically significantly different from zero in the population ($H_0: \rho = 0$).

We can perform this test using `cor.test()`:

```{r}
cor.test(d$x, d$y)
```

### Interpreting the Output

1. **$t$-statistic and degrees of freedom**: $t(3) = 3.55$, where $df = n - 2 = 5 - 2 = 3$.
2. **$p$-value**: $p = 0.038$, which is below the threshold of $\alpha = .05$. We reject the null hypothesis and conclude that $x$ and $y$ are significantly positively correlated.
3. **95% Confidence Interval**: $[0.11, 0.99]$. Since the interval does not include $0$, it further confirms statistical significance.

### APA-Style Reporting

> *"A Pearson correlation revealed a strong, positive, and statistically significant relationship between the two variables, $r(3) = .90$, $p = .038$, 95% CI $[0.11, 0.99]$."*

### The test behind `cor.test()`

Under the hood, `cor.test()` performs a **t-test** on the correlation coefficient: it computes $t = r \sqrt{n - 2} / \sqrt{1 - r^2}$ and compares it to a t-distribution with $n - 2$ degrees of freedom. So testing a correlation uses the same logic as the t-tests you saw in [Comparing two Groups](5.01-comparing-two-groups.qmd).
