---
title: "Grouping and summarizing"
subtitle: "Calculating and interpreting summary statistics."
---

::: {.callout-note title="Where you are in your learning journey" icon=false}
**Where you come from.** You arrive here from [Data Frames](data-frames.qmd). You can load a data frame and pick the columns and rows you need.

**What this module adds.** You'll group data and summarise it per group with `group_by()` and `summarise()`.

**Why it is valuable.** Comparing groups is at the heart of a lot of research. This is also the stepping stone to descriptive statistics and plotting.

**Where it leads.** Next up: [Pivoting and joining](pivoting-and-joining.qmd), [Descriptive Statistics](descriptive-statistics.qmd) and [Data Visualization](data-visualization.qmd).
:::

# Data Summarization

The functions used in earlier parts on data preparation worked on individual rows.
Sometimes, you need to compute properties of groups of rows (cases).
This is called aggregation (or summarization) and in the tidyverse uses the `group_by` function
followed by either `summarize` or `mutate`.

::: {.callout-note title="Required packages and data"}
```{r, message=F, warning=F}
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
d <- read_csv(url) |>
  select(-URL) |>
  rename(Rep = `Republican Support`, Dem = `Democratic Support`)
d
```
:::

# Grouping rows

We can use the `group_by` function to group data by a specific column, for example by `Question`:

```{r}
d |>
  group_by(Question)
```

The data itself hasn't changed its values, but dplyr records that the data is now grouped by `Question` (with 8 groups in total).

# Summarizing

To summarize grouped data, follow `group_by` with a call to `summarize`.
The syntax is `summarize(column = calculation, ...)`.
Crucially, the calculation must compute a single summary value given a vector of values.
Very common summarization functions are `mean()`, `sd()`, `median()`, and `n()`.

For example, the following computes the average support per question:

```{r}
d |>
  group_by(Question) |>
  summarize(Support = mean(Support, na.rm = TRUE)) |>
  arrange(desc(Support))
```

`summarize` reduces the rows equal to the number of groups (8), keeping only the grouping variables and the newly calculated summary columns.

You can also compute summaries of multiple values:

```{r}
d |>
  group_by(Question) |>
  summarize(
    Dem = mean(Dem, na.rm = TRUE),
    Rep = mean(Rep, na.rm = TRUE),
    diff = mean(Dem - Rep, na.rm = TRUE)
  ) |>
  arrange(desc(diff))
```

Another essential helper is `n()` (without arguments), which counts the number of rows in each group:

```{r}
d |>
  group_by(Question) |>
  summarize(
    n = n(),
    mean = mean(Support, na.rm = TRUE),
    sd = sd(Support, na.rm = TRUE)
  )
```

# Using mutate with group_by

While `summarize` collapses multiple rows into a single summary row, using `mutate` after `group_by` preserves all rows and adds the group-level summary as a new column:

```{r}
d2 <- d |>
  group_by(Question) |>
  mutate(
    avg_support = mean(Support, na.rm = TRUE),
    diff = Support - avg_support
  )
d2
```

# Ungrouping

Finally, you can use `ungroup()` to remove grouping metadata. When you are done computing grouped operations, calling `ungroup()` ensures subsequent functions treat the data frame as a single whole:

```{r}
d2 |>
  ungroup() |>
  summarize(overall_sd = sd(diff, na.rm = TRUE))
```
