---
title: "String processing"
subtitle: "String manipulation with stringr: regex detection, extraction, and trimming."
---

::: {.callout-note title="Where you are in your learning journey" icon=false}
**Where you come from.** You arrive here from [Data Frames](data-frames.qmd). You can load a data frame and work with its columns.

**What this module adds.** You'll work with text columns using `stringr`: detecting, extracting, replacing, and cleaning text patterns.

**Why it is valuable.** Text is everywhere in communication data (open answers, names, social media posts) and cleaning text is the first step to analysing it.

**Where it leads.** Next up: [Text as Data](text-as-data.qmd).
:::

# Introduction

In communication science, much of the data we work with contains text: survey open-ended responses, headlines, social media posts, or category codes that need standardization.
Before conducting computational content analysis, we often need basic tools to inspect, clean, and manipulate character strings.

The `stringr` package (part of the core tidyverse) provides a consistent and friendly set of functions for string manipulation, where all functions start with `str_`.

::: {.callout-note title="Required Packages"}
```{r, message=FALSE}
library(tidyverse)
```
:::

# Basic String Operations

Let's create a small example vector of news headlines:

```{r}
headlines <- c(
  "Breaking: Prime minister announces new climate policy!",
  "Economy shows modest growth in Q3",
  "Breaking: New survey reveals rising public concern over inflation",
  "Local election debate ends in stalemate"
)
headlines
```

## Detecting Patterns: `str_detect()`

`str_detect()` tests whether a string matches a pattern, returning `TRUE` or `FALSE`. It is frequently used inside `filter()`:

```{r}
str_detect(headlines, "Breaking")
```

To make searches case-insensitive, wrap the pattern in `regex(..., ignore_case = TRUE)`:

```{r}
str_detect(headlines, regex("election", ignore_case = TRUE))
```

## Extracting Patterns: `str_extract()`

`str_extract()` pulls out the first matching pattern:

```{r}
str_extract(headlines, "Q[1-4]")
```

## Replacing Patterns: `str_replace()` and `str_replace_all()`

`str_replace()` replaces the first match, while `str_replace_all()` replaces every occurrence:

```{r}
str_replace(headlines, "Breaking: ", "")
```

## Modifying Case and Whitespace

Common data cleaning tasks involve standardizing casing and trimming unnecessary whitespace:

```{r}
dirty_text <- c("  United Nations ", "EU ", " climate change  ")

str_trim(dirty_text)
str_to_lower(dirty_text)
```

# Using `stringr` inside Data Frames

In practice, string operations are most powerful inside `mutate()` and `filter()`:

```{r}
df <- tibble(
  title = headlines,
  source = c("Daily News", "Financial Times", "Daily News", "Metro")
)

df |> 
  filter(str_detect(title, "Breaking")) |> 
  mutate(clean_title = str_replace(title, "Breaking: ", ""))
```

> For full details on regular expressions and advanced string operations, see [R for Communication Science](https://r4css.vanatteveldt.com/).
