---
title: Function Documentation
---

# Viewing Function Documentation in R

One of the things that can new users in R can find overwhelming, is that they think they need to learn all functions by heart.
This is not the case!
Aside from a handfull of functions that you will use all the time, you will often need to look up how to use a function.
Rather than learning everything by heart, you therefore need to learn some tricks for how to quickly look up information about functions.

One of the most important tricks is to use the built-in help system in R.
You can quickly access documentation for any function using the `?` symbol. 
This is a powerful tool that can help you understand how to use functions, what arguments they require, and what they return.

## How to Use the `?` Symbol

To view the documentation for a specific function, you simply need to type `?` followed by the function name. For example, if you want to learn more about the `mean()` function, you would type:

```{r, eval=F}
?mean
```

This will open the help page, often in the bottom right pane of RStudio.

## How to read the help page

The help page for a function is divided into several sections. 
The most important sections are:

### Description

A brief description of what the function does.
For the `mean()` function, the description is: *Generic Function for the (Arithmetic) Mean*.

By generic function, they mean that the function can have multiple implementations.
When you think of the mean, you are probably thinking of the mean of a vector of numbers.

```{r}
x = c(1,2,3,4)
mean(x)
```

But you can do more!
For example, you can also calculate the mean of a vector of `Date` values:

```{r}
dates = as.Date(c("2021-01-01", "2021-01-03"))
mean(dates)
```


### Usage

The syntax of the function, including all the arguments it takes.
For example, the `mean()` function has the following usage:

```{r, eval=F}
mean(x, ...)

## Default S3 method:
mean(x, trim = 0, na.rm = FALSE, ...)
```

The first part tells you that the most basic way to use the function is to provide an argument called `x`. What `x` is, is explained in the *arguments* section that we discuss below.

The `...` at the end means that the function can take additional arguments. 
This is because the mean function is a *generic function*.
Depending on the type of input you provide (e.g., numbers, dates), some arguments might not be relevant.

The second part tells you that the default method for the mean function has two additional arguments in addition to `x`: `trim`, and `na.rm`.
Note an important difference with the `x` argument: these arguments have default values (0 and FALSE, respectively).
This means that these arguments are **optional**.
If you don't specify them, the function will use these default values.

For example, notice that the `na.rm` argument is set to `FALSE` by default. 
As we can see in the Arguments section, this means that the function will not remove missing values by default.
(`NA` stands for `Not Available`, and is used in R to indicate missing values, so `na.rm` is short for `remove NAs`).

```{r}
x_with_missing <- c(1, 2, 3, NA, 4)
mean(x_with_missing)
```

If we want to remove missing values, we can set `na.rm` to `TRUE`:

```{r}
mean(x_with_missing, na.rm=TRUE)
```

::: {.callout-note title="To name or not to name your arguments" collapse="true"}

Notice how in the code above we specify the argument name `na.rm = TRUE` to indicate that we want to use this optional argument.
For the `x` argument we don't need to specify the argument name, because it's the first argument and the function knows that the first argument is `x`.
Generally speaking, if you don't specify the argument name, R will assume that you are providing the arguments in the order that they are listed in the usage section.
Let's think a bit about when we should and should not use argument names!

You *could* decide to always use argument names:

```{r}
mean(x=x_with_missing, na.rm=TRUE)
```

This is fine, and sometimes you might want to do this for sake of clarity.
But it's also often unnecessary.
For the `mean` function, it is obvious that the first argument is the input over which you want to calculate the mean, so you don't need to specify the argument name.

On the opposite end of the spectrum, you could decide to never use argument names:

```{r}
mean(x_with_missing, 0, TRUE)
```

Here the three arguments follow the order in the usage section: `x`, `trim`, `na.rm`.

This has two obvious downsides:

* It is not obvious what the `0` and `TRUE` arguments are. The reader might thus have to look up the function documentation.
* We now also need to specify the `trim` argument, because it comes before `na.rm` in the usage section.

So in general, it is often good to use argument names for optional arguments, like `na.rm`.
For required arguments, like `x`, it is often not necessary.
Arguably the *best* way to use the `mean` function with `na.rm` is therefore:

```{r}
mean(x_with_missing, na.rm=TRUE)
```

:::

### Arguments

A description of all the arguments that the function takes.
This should cover all the arguments that are listed in the usage section.

For example, the `mean()` function explains that the `x` argument is can be a numeric vector, but also something like a `logical` or `date` vector.
For the `na.rm` argument it explains that if set to `TRUE`, missing values will be removed before calculating the mean.

### Value

The value section explains what the function returns (i.e. the output).

### Examples

The examples section shows you how to use the function.
Honestly, this is often the most useful part of the help page.
If you are not sure how to use a function, a great way to learn is to look at the examples.
Usually, you can directly copy-paste these examples into your script and run them to see how the function works.


# Using tab completion


A very usefull trick in RStudio, and in programming in general, is to use tab completion.
The tab key on your keyboard (the one above the caps lock key) can be used to complete the name of a function or variable.
And if there are multiple ways in which the name can be completed, R will show you all the options.
(On some devices you might need to press `Tab` twice, or it might not work at all).

For example, type the following code in your R script:

```{r, eval=F}
mean()
```

Now place your cursor between the parentheses and press the `Tab` key.
This should now list all the arguments for the `mean` function, including their descriptions!

Since the mean function is empty, you should only have seen the `x` argument (and the `...` argument).
But you can also use tab completion when adding additional arguments.
Type the following code in your R script:

```{r, eval=F}
x = c(1,2,3,4)
mean(x, )
```

Now place your cursor between the parentheses after the comma, and press the `Tab` key.
Now you should see the `trim` and `na.rm` arguments, because RStudio knows that these arguments apply to a `mean` function with a numeric input!

## Try using tab completion everywhere

Well ok, not everywhere. But you might be surprised how often it can help you.
It can even help you find files on your computer.
If you use tab completion between quotes, RStudio will show you all the files in your working directory that match the characters you've typed so far.
So you can use this inside functions like `read_csv` to quickly find the file you want to read.

```{r, eval=F}  
library(tidyverse)
read_csv("")
```

Try it out!